---
title: "exploring decontaminated metabarcoding reads from DBO samples"
author: "Kimberly Ledger"
date: "2024-10-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

libraries
```{r}
library(tidyverse)
library(ggplot2)
rename <- dplyr::rename
library(viridis)
library(ggrepel)
```

the data 
```{r}
taxon_table <- read.csv("/home/kimberly.ledger/dbo_metabarcoding/outputs/decontaminated_taxon_table_filtered.csv") %>%
  select(!X)  %>%
  filter(sample_type == "sample")

metadata <- read.csv("/home/kimberly.ledger/dbo_metabarcoding/data/NBS_SBS_DBO_metadata.csv") %>%
  filter(project == "DBO")

metadata_mini <- metadata %>%
  select(extraction_ID, alternative_ID, sample_type, project, collection_year, location1, location2, depth, longitude, latitude) %>%
  unite("project_year", project, collection_year, sep = "_", remove = F) %>%
  mutate(extraction_ID = ifelse(project_year == "DBO_2021", alternative_ID, extraction_ID)) %>% # for DBO2021 samples, replace extraction_ID with alternative_ID
  select(!alternative_ID)
```

get additional cruise metadata - this is not a final version for 2023 samples 
```{r}
cruise_meta <- read.csv("/home/kimberly.ledger/dbo_metabarcoding/data/SikuliaqDBO_2021_2023_cruisedata_20241015.csv") %>%
  mutate(temp_C = (temperature_ch1 + temperature_ch2)/2) %>%
  mutate(salinity = (salinity_ch1 + salinity_ch2)/2)
```

```{r}
metadata_join <- metadata_mini %>%
  left_join(cruise_meta) %>%
  mutate(temp_C = ifelse(extraction_ID == "e04213", -0.89975, temp_C),   ## filling in missing data by taking the average on the station before and after at same depth 
         temp_C = ifelse(extraction_ID == "e04214", 2.77975, temp_C),
         salinity = ifelse(extraction_ID == "e04213", 32.45635, salinity),
         salinity = ifelse(extraction_ID == "e04214", 30.39102, salinity),
         Collection_Time_local = ifelse(extraction_ID == "E1090.SKQ2021", '10:31', Collection_Time_local),
         Collection_Time_local = ifelse(extraction_ID == "E1088.SKQ2021", '10:31', Collection_Time_local),
         Collection_Time_local = ifelse(extraction_ID == "E1089.SKQ2021", '10:31', Collection_Time_local))
```


since processing of the taxon table included samples from NBS and SBS surveys, let me go ahead are remove any taxa that were found in those but NOT in any DBO samples 
```{r}
dbo_taxa <- taxon_table %>%
  group_by(taxon) %>%
  summarize(taxon_reads = sum(tot_reads)) %>%
  filter(taxon_reads > 55)  ## rare taxa (in terms of total read count) seem to cause problems with community analyses, removing them here

taxon_table <- taxon_table %>%
  filter(taxon %in% dbo_taxa$taxon) %>%
  filter(class == "Actinopteri")                 ### remove marine mammals! 

#also will remove any taxa that only pop up in a single extraction across all DBO samples
taxa_in_one_extract <- taxon_table %>%
  filter(tot_reads > 0) %>%
  group_by(taxon) %>%
  summarize(num = n()) %>%
  filter(num == 1)

taxon_table <- taxon_table %>%
  filter(!taxon %in% taxa_in_one_extract$taxon)

#check number of reads per extraction - with these additional filters some may have dropped 
taxon_table %>%
  group_by(extraction_ID) %>%
  summarize(reads = sum(tot_reads)) %>%
  arrange(reads)

#remove the extraction that now had no reads
taxon_table <- taxon_table %>%
   filter(extraction_ID != "E1066.SKQ2021")

#remove the high latitude station (CK9) from 2023 with VERY DEEP samples - might want to add back in to dataset, but just want to make sure it's not responsible for driving lat or depth patterns 
taxon_table <- taxon_table %>%
  filter(extraction_ID != "e04171") %>%
  filter(extraction_ID != "e04172") %>%
  filter(extraction_ID != "e04173") %>%
  filter(extraction_ID != "e04175")

#check the number of taxa per extraction
taxon_table %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  summarize(n_taxa = n()) %>%
  arrange(n_taxa)
```

set up two types of distance matrices
1) binary (presence/absence)
2) semiquantitative using a eDNA index (Kelly et al. 2019)

let me work with just the minimum info for now - remove extraction ID's with no reads 
```{r}
taxa_df <- taxon_table[,c(1,6,13)] %>%
  filter(!is.na(tot_reads)) %>%
  rename(reads = tot_reads)
```

calculate eDNA index according to Kelly et al. 2019
```{r}
library(vegan)

index_df <- taxa_df %>%
    group_by(extraction_ID) %>%
    mutate(Tot = sum(reads),
              Prop = reads / Tot) %>% ## this creates the proportion on each bottle replicate
    select(-reads, -Tot) %>%
   pivot_wider(names_from = taxon, values_from = Prop)
  
ids <- index_df$extraction_ID
index_df <- index_df[,-1]

wis_index <- wisconsin(index_df)

rowSums(wis_index)
wis_index$extraction_ID <- ids
```


let me compare tables 
```{r}
reads_df_wide <- taxa_df %>%
  pivot_wider(names_from = taxon, values_from = reads)

rowSums(reads_df_wide[,-1])
colSums(reads_df_wide[,-1])
dim(reads_df_wide[,-1])
any(duplicated(reads_df_wide[,-1]))


binary_df_wide <- taxa_df %>%
  mutate(pres = ifelse(reads > 0, 1, 0)) %>%
  select(!reads) %>%
  pivot_wider(names_from = taxon, values_from = pres)

any(duplicated(binary_df_wide[,-1]))

index_df_wide <- wis_index %>%
  select(extraction_ID, everything())
```

get some metadata to run multivariate analyses 
for now, i'm going to just work with field samples (i.e. ignore any reads in remaining field negatives, etc)
```{r}
meta_for_nmds <- metadata_join  %>%
  filter(sample_type == "sample") %>%
  select(extraction_ID, collection_year, location1, location2, depth, longitude, latitude, Collection_Time_local, temp_C, salinity) %>%
  mutate(depth = as.numeric(depth)) %>%
  mutate(depth_cat = ifelse(depth <= 11, "10m", NA),
         depth_cat = ifelse(depth == 24.2, "30m", depth_cat),    ## binning
         depth_cat = ifelse(depth == 30, "30m", depth_cat),
         depth_cat = ifelse(depth > 30, "bottom", depth_cat)) %>%
  filter(extraction_ID %in% reads_df_wide$extraction_ID)
```


run an NMDS - getting errors for possibly having insufficient data... 
```{r}
## fill in 0's for the extractions with 0 reads - maybe i actually want to just remove them? 
#reads_df_wide[is.na(reads_df_wide)] <- 0 
#index_df_wide[is.na(index_df_wide)] <- 0 
#binary_df_wide[is.na(binary_df_wide)] <- 0 

#binary_jaccard <- vegdist(binary_df_wide[,-1], method = "bray")
#mds <- metaMDS(binary_jaccard)

binary_NMS_2<-metaMDS(binary_df_wide[,-1], distance = "jaccard", k=2, try=200, autotransform = FALSE, maxit=1000)
binary_NMS_3<-metaMDS(binary_df_wide[,-1], distance = "jaccard", k=3, try=200, autotransform = FALSE, maxit=1000)
binary_NMS_4<-metaMDS(binary_df_wide[,-1], distance = "jaccard", k=4, try=200, autotransform = FALSE, maxit=1000)
binary_NMS_5<-metaMDS(binary_df_wide[,-1], distance = "jaccard", k=5, try=200, autotransform = FALSE, maxit=1000)
binary_NMS_6<-metaMDS(binary_df_wide[,-1], distance = "jaccard", k=6, try=200, autotransform = FALSE, maxit=1000)
binary_NMS_7<-metaMDS(binary_df_wide[,-1], distance = "jaccard", k=7, try=200, autotransform = FALSE, maxit=1000)
binary_NMS_8<-metaMDS(binary_df_wide[,-1], distance = "jaccard", k=8, try=200, autotransform = FALSE, maxit=1000)
binary_NMS_9<-metaMDS(binary_df_wide[,-1], distance = "jaccard", k=9, try=200, autotransform = FALSE, maxit=1000)
binary_NMS_10<-metaMDS(binary_df_wide[,-1], distance = "jaccard", k=10, try=200, autotransform = FALSE, maxit=1000)
#binary_NMS_2$stress
#stressplot(binary_NMS)

stress_2d <- binary_NMS_2$stress
stress_3d <- binary_NMS_3$stress
stress_4d <- binary_NMS_4$stress
stress_5d <- binary_NMS_5$stress
stress_6d <- binary_NMS_6$stress
stress_7d <- binary_NMS_7$stress
stress_8d <- binary_NMS_8$stress
stress_9d <- binary_NMS_9$stress
stress_10d <- binary_NMS_10$stress

k_values <- c(2, 3, 4, 5, 6, 7, 8, 9, 10)
stress_values <- c(stress_2d, stress_3d, stress_4d, stress_5d, stress_6d, stress_7d, stress_8d, stress_9d, stress_10d)

plot(k_values, stress_values, type = "b", xlab = "Number of Dimensions (k)", ylab = "Stress Value", main = "Choosing Optimal Dimensions for NMDS")

#index_NMS<-metaMDS(index_df_wide[,-1], distance = "bray", k=5, try=200, autotransform = FALSE, maxit=1000)

index_NMS_2<-metaMDS(index_df_wide[,-1], distance = "bray", k=2, try=200, autotransform = FALSE, maxit=1000)
index_NMS_3<-metaMDS(index_df_wide[,-1], distance = "bray", k=3, try=200, autotransform = FALSE, maxit=1000)
index_NMS_4<-metaMDS(index_df_wide[,-1], distance = "bray", k=4, try=200, autotransform = FALSE, maxit=1000)
index_NMS_5<-metaMDS(index_df_wide[,-1], distance = "bray", k=5, try=200, autotransform = FALSE, maxit=1000)
index_NMS_6<-metaMDS(index_df_wide[,-1], distance = "bray", k=6, try=200, autotransform = FALSE, maxit=1000)
index_NMS_7<-metaMDS(index_df_wide[,-1], distance = "bray", k=7, try=200, autotransform = FALSE, maxit=1000)
index_NMS_8<-metaMDS(index_df_wide[,-1], distance = "bray", k=8, try=200, autotransform = FALSE, maxit=1000)
index_NMS_9<-metaMDS(index_df_wide[,-1], distance = "bray", k=9, try=200, autotransform = FALSE, maxit=1000)
index_NMS_10<-metaMDS(index_df_wide[,-1], distance = "bray", k=10, try=200, autotransform = FALSE, maxit=1000)

stress_2d <- index_NMS_2$stress
stress_3d <- index_NMS_3$stress
stress_4d <- index_NMS_4$stress
stress_5d <- index_NMS_5$stress
stress_6d <- index_NMS_6$stress
stress_7d <- index_NMS_7$stress
stress_8d <- index_NMS_8$stress
stress_9d <- index_NMS_9$stress
stress_10d <- index_NMS_10$stress

k_values <- c(2, 3, 4, 5, 6, 7, 8, 9, 10)
stress_values <- c(stress_2d, stress_3d, stress_4d, stress_5d, stress_6d, stress_7d, stress_8d, stress_9d, stress_10d)

plot(k_values, stress_values, type = "b", xlab = "Number of Dimensions (k)", ylab = "Stress Value", main = "Choosing Optimal Dimensions for NMDS")

#index_NMS$stress
#stressplot(index_NMS)

#reads_NMS<-metaMDS(reads_df_wide[,-1], distance = "bray", k=2, try=100, autotransform = TRUE, maxit=1000)
```

set the nmds dataset i want to plot 
```{r}
#NMS_data <- binary_NMS_6
NMS_data <- index_NMS_6
#NMS_data <- reads_NMS #not going to explore this one, nmds value are very very small
```


```{r}
#create vectors with the NMS attributes
NMS_coordinates<-scores(NMS_data,display="sites")
NMS_axes<-as.data.frame(NMS_coordinates)
NMS_scores<-scores(NMS_data,display="species")
```

link back sample metadata
```{r}
for_ploting<-as.data.frame(cbind(NMS_coordinates,meta_for_nmds))
```

```{r}
#plot the stress
stressplot(NMS_data)
NMS_data$stress
```

plot - color by latitude 
```{r}
nmds.plot <- ggplot(for_ploting, aes(x=NMDS1, y=NMDS2))+ #sets up the plot
  geom_point(aes(NMDS1, NMDS2, colour = latitude, shape = factor(collection_year)), size = 2)+ #adds site points to plot, shape determined by year, colour determined by location
  coord_fixed()+
  theme_classic()+ 
  theme(panel.background = element_rect(fill = NA, colour = "black", size = 1, linetype = "solid"))+
  labs(colour = "Latitude", shape = "Year", title = "by latitude") + # add legend labels for Station
  theme(legend.position = "right", 
        legend.text = element_text(size = 12), 
        legend.title = element_text(size = 12), 
        axis.text = element_text(size = 10)) #+ # add legend at right of plot
 #binary
   #xlim(-0.035,0.02) +
   #ylim(-0.04,0.016)
 #index  
  #xlim(-0.022, 0.011) +
  #ylim(-0.019,0.008)

nmds.plot
```
now in this iteration, E1108.SKQ2021 stands out as an outlier... 

plot - color by longitude 
```{r}
nmds.plot <- ggplot(for_ploting, aes(x=NMDS1, y=NMDS2))+ #sets up the plot
  geom_point(aes(NMDS1, NMDS2, colour = longitude, shape = factor(collection_year)), size = 2)+ #adds site points to plot, shape determined by year, colour determined by location
  coord_fixed()+
  theme_classic()+ 
  theme(panel.background = element_rect(fill = NA, colour = "black", size = 1, linetype = "solid"))+
  labs(colour = "Longitude", shape = "Year", title = "by longitude") + # add legend labels for Station
  theme(legend.position = "right", 
        legend.text = element_text(size = 12), 
        legend.title = element_text(size = 12), 
        axis.text = element_text(size = 10)) #+ # add legend at right of plot
 #binary
   #xlim(-0.035,0.02) +
   #ylim(-0.04,0.016)
 #index  
  #xlim(-0.022, 0.011) +
  #ylim(-0.019,0.008)

nmds.plot
```

plot - color by transect 
```{r}
nmds.plot <- ggplot(for_ploting, aes(x=NMDS1, y=NMDS2))+ #sets up the plot
  geom_point(aes(NMDS1, NMDS2, colour = factor(location2), shape = factor(collection_year)), size = 2)+ #adds site points to plot, shape determined by year, colour determined by location
  coord_fixed()+
  theme_classic()+ 
  theme(panel.background = element_rect(fill = NA, colour = "black", size = 1, linetype = "solid"))+
  labs(colour = "Station", shape = "Year", title = "by transect/station") + # add legend labels for Station
  theme(legend.position = "right", 
        legend.text = element_text(size = 12), 
        legend.title = element_text(size = 12), 
        axis.text = element_text(size = 10)) #+ # add legend at right of plot
 #binary
   #xlim(-0.035,0.02) +
   #ylim(-0.04,0.016)
 #index  
  #xlim(-0.022, 0.011) +
  #ylim(-0.019,0.008)

nmds.plot
```

plot - color by depth
```{r}
nmds.plot <- ggplot(for_ploting, aes(x=NMDS1, y=NMDS2))+ #sets up the plot
  geom_point(aes(NMDS1, NMDS2, colour = factor(depth_cat), shape = factor(collection_year)), size = 2)+ #adds site points to plot, shape determined by year, colour determined by location
  coord_fixed()+
  theme_classic()+ 
  theme(panel.background = element_rect(fill = NA, colour = "black", size = 1, linetype = "solid"))+
  labs(colour = "Depth", shape = "Year", title = "by depth") + # add legend labels for Station
  theme(legend.position = "right", 
        legend.text = element_text(size = 12), 
        legend.title = element_text(size = 12), 
        axis.text = element_text(size = 10)) #+ # add legend at right of plot
 #binary
   #xlim(-0.035,0.02) +
   #ylim(-0.04,0.016)
 #index  
  #xlim(-0.022, 0.011) +
  #ylim(-0.019,0.008)

nmds.plot
```

plot - color by year 
```{r}
nmds.plot <- ggplot(for_ploting, aes(x=NMDS1, y=NMDS2))+ #sets up the plot
  geom_point(aes(NMDS1, NMDS2, colour = factor(collection_year), shape = factor(collection_year)), size = 2)+ #adds site points to plot, shape determined by year, colour determined by year
  coord_fixed()+
  theme_classic()+ 
  theme(panel.background = element_rect(fill = NA, colour = "black", size = 1, linetype = "solid"))+
  labs(colour = "Year", shape = "Year", title = "by year") + # add legend labels for Station
  theme(legend.position = "right", 
        legend.text = element_text(size = 12), 
        legend.title = element_text(size = 12), 
        axis.text = element_text(size = 10)) #+ # add legend at right of plot
  #binary
   #xlim(-0.035,0.02) +
   #ylim(-0.04,0.016)
 #index  
  #xlim(-0.022, 0.011) +
  #ylim(-0.019,0.008)

nmds.plot
```

```{r}
#spp.fit <- envfit(NMS_data, binary_df_wide[,-1], permutations = 999) # this fits species vectors
spp.fit <- envfit(NMS_data, index_df_wide[,-1], permutations = 999) # this fits species vectors

spp.scrs <- as.data.frame(scores(spp.fit, display = "vectors")) #save species intrinsic values into dataframe
spp.scrs <- cbind(spp.scrs, Species = rownames(spp.scrs)) #add species names to dataframe
spp.scrs <- cbind(spp.scrs, pval = spp.fit$vectors$pvals) #add pvalues to dataframe so you can select species which are significant
#spp.scrs<- cbind(spp.scrs, abrev = abbreviate(spp.scrs$Species, minlength = 6)) #abbreviate species names
sig.spp.scrs <- subset(spp.scrs, pval<=0.05) #subset data to show species significant at 0.05

head(spp.scrs)

library(ggrepel)
# nmds.plot +
#   geom_segment(data = sig.spp.scrs, aes(x = 0, xend=NMDS1, y=0, yend=NMDS2), arrow = arrow(length = unit(0.25, "cm")), colour = "grey10", lwd=0.3) + #add vector arrows of significant species
#   ggrepel::geom_text_repel(data = sig.spp.scrs, aes(x=NMDS1, y=NMDS2, label = Species), cex = 3, direction = "both", segment.size = 0.25)


#plot just significant species 
nmds.plot.spp <- ggplot() + 
  geom_point(data = sig.spp.scrs, aes(NMDS1, NMDS2, colour = Species)) + 
  ggrepel::geom_text_repel(data = sig.spp.scrs, aes(x=NMDS1, y=NMDS2, label = Species), cex = 3, direction = "both", segment.size = 0.25) + 
  theme_classic()+ 
  theme(legend.position = "none") +
  labs(title = "significant species")

nmds.plot.spp   
```

nmds1 - latitude
nmds2 - year 

## investigate the outlier samples 
```{r}
taxon_table %>%
  filter(extraction_ID == "E1108.SKQ2021") %>%
  filter(tot_reads > 0)
```

```{r}
taxon_table_meta %>% 
  filter(taxon == "Gadus macrocephalus") %>%
  filter(tot_reads > 0)
```

notes from previous runs: 
Allosmerus elongatus in four samples - each with only 2 or 3 reads 
Icelus only in one sample

remove all species with only one (or few detections?)
remove taxa with very very few reads (ie. Allosmerus elongatus)

## test for differences in communities between lat/long/depth/year using PERMANOVA 

first which variable are correlated? 
```{r}
plot(meta_for_nmds$latitude ~ meta_for_nmds$longitude)
```

strong relationship between lat and long, so just going to test latitude 

```{r}
plot(meta_for_nmds$latitude ~ meta_for_nmds$depth)
```

```{r}
boxplot(meta_for_nmds$latitude ~ meta_for_nmds$depth_cat)
```

```{r}
boxplot(meta_for_nmds$latitude ~ meta_for_nmds$collection_year)
```

```{r}
boxplot(meta_for_nmds$latitude ~ as.factor(meta_for_nmds$location2))
```

```{r}
plot(meta_for_nmds$latitude ~ meta_for_nmds$temp_C, col = meta_for_nmds$collection_year)
```

```{r}
plot(meta_for_nmds$depth ~ meta_for_nmds$temp_C, col = meta_for_nmds$collection_year)
```

```{r}
boxplot(meta_for_nmds$temp_C ~ meta_for_nmds$collection_year)
```

```{r}
plot(meta_for_nmds$latitude ~ meta_for_nmds$salinity, col = meta_for_nmds$collection_year)
```

```{r}
plot(meta_for_nmds$depth ~ meta_for_nmds$salinity, col = meta_for_nmds$collection_year)
```

```{r}
boxplot(meta_for_nmds$salinity ~ meta_for_nmds$collection_year)
```

```{r}
meta_for_nmds$time_of_day_posix <- as.POSIXct(meta_for_nmds$Collection_Time_local, format = "%H:%M", tz = "UTC")
plot(meta_for_nmds$latitude ~ meta_for_nmds$time_of_day_posix, col = meta_for_nmds$collection_year)
```


```{r}
binary_dist <- vegdist(binary_df_wide[,-1], method = "jaccard")

# Run PERMANOVA with latitude as the predictor
binary_lat <- adonis2(binary_dist ~ meta_for_nmds$latitude, permutations = 999)
binary_lat
binary_depth <- adonis2(binary_dist ~ meta_for_nmds$depth, permutations = 999)      ## when include very deep samples, consider log() transformation
binary_depth
binary_depth_cat <- adonis2(binary_dist ~ meta_for_nmds$depth_cat, permutations = 999)
binary_depth_cat
binary_year <- adonis2(binary_dist ~ meta_for_nmds$collection_year, permutations = 999)
binary_year
binary_temp <- adonis2(binary_dist ~ meta_for_nmds$temp_C, permutations = 999)
binary_temp
binary_salinity <- adonis2(binary_dist ~ meta_for_nmds$salinity, permutations = 999)
binary_salinity

# Convert the time to POSIXct (or POSIXlt)
time_posix <- strptime(meta_for_nmds$Collection_Time_local, format = "%H:%M")
# Convert the time to numeric hours (0 to 24)
meta_for_nmds$numeric_hours <- as.numeric(format(time_posix, "%H")) + as.numeric(format(time_posix, "%M")) / 60

binary_tod <- adonis2(binary_dist ~ meta_for_nmds$numeric_hours, permutations = 999)
binary_tod
```

for binary format, latitude and year are significant (and now also temp and salinity)

```{r}
index_dist <- vegdist(index_df_wide[,-1], distance = "bray")

# Run PERMANOVA with latitude as the predictor
index_lat <- adonis2(index_dist ~ meta_for_nmds$latitude, permutations = 999)
index_lat
index_depth <- adonis2(index_dist ~ meta_for_nmds$depth, permutations = 999) ## when include very deep samples, consider log() transformation
index_depth
index_depth_cat <- adonis2(index_dist ~ meta_for_nmds$depth_cat, permutations = 999)
index_depth_cat
index_year <- adonis2(index_dist ~ meta_for_nmds$collection_year, permutations = 999)
index_year
index_temp <- adonis2(index_dist ~ meta_for_nmds$temp_C, permutations = 999)
index_temp
index_salinity <- adonis2(index_dist ~ meta_for_nmds$salinity, permutations = 999)
index_salinity
index_tod <- adonis2(index_dist ~ meta_for_nmds$numeric_hours, permutations = 999)
index_tod
```

significant continuous: latitude, depth (and now temp and salinity)
significant categorical: year

## conduct PERMDISP (Permutational Multivariate Dispersion Test) for significant PERMANOVA results
- only for CATEGORICAL PREDICTORS 
```{r}
# Assuming you have already computed a distance matrix (e.g., Bray-Curtis dissimilarity) - going to use 'index_dist'

# Step 1: Use betadisper to calculate the dispersion of groups in principal coordinate space

dispersion <- betadisper(index_dist, as.factor(meta_for_nmds$collection_year))
#dispersion <- betadisper(binary_dist, as.factor(meta_for_nmds$collection_year))

# Step 2: Perform a permutation test for multivariate homogeneity of group dispersions (PERMDISP)
dispersion_test <- permutest(dispersion, permutations = 999)

# Step 3: View the results of the test
print(dispersion_test)

# Step 4: Optionally, visualize the distances to group centroids
plot(dispersion)
boxplot(dispersion, main = "Group Dispersion by Year")
```

okay, so for collection year, since the PERMDISP is also significant, this suggests that the differences in community structure may be partially driven by unequal dispersion of samples in the multivariate space, not just differences in centroid positions - will need to be caution about interpretation of the PERMANOVA, as differences in comp between years might be confounded by heterogeneity in dispersion 

hmm, even though PERMDISP is significant, the plot make it seem like dispersion is pretty similar between years....


let me try a distance-based redundancy analysis (dbRDA) for continuous variables - latitude and depth
```{r}
# Fit the dbRDA model - using capscale() - constrained analysis of principal coordinates

rda_index <- capscale(index_df_wide[,-1] ~ meta_for_nmds$latitude + meta_for_nmds$depth + as.factor(meta_for_nmds$collection_year) + meta_for_nmds$temp_C + meta_for_nmds$salinity, distance = "bray")

#rda_index <- capscale(binary_df_wide[,-1] ~ meta_for_nmds$latitude + meta_for_nmds$depth + as.factor(meta_for_nmds$collection_year), distance = "jaccard")  ## need to change variable name at somepoint to avoid confusion...

anova_terms <- anova(rda_index, by = "terms", permutations = 999)
anova_terms 

summary(rda_index)

# Assuming capscale_model is already defined
eigenvalues <- rda_index$CA$eig
total_inertia <- sum(eigenvalues)

# Calculate proportion explained for CAP1 and CAP2
proportion_variance <- eigenvalues / total_inertia * 100

# First use the regular plot option to get a feel for the data
plot(rda_index, scaling = 2, choices = c(1,2))
plot(rda_index, scaling = 2, choices = c(1,3))

# CAP1 = latitude, CAP2 = year, CAP3 = depth (CAP3 changes when adding in temp and salinity to the model)

# Extract site and species scores
site_scores <- as.data.frame(vegan::scores(rda_index, display = "sites", choices = 1:3))
#site_scores <- as.data.frame(rda_index$CCA$u[, 1:3])
site_scores$Sample <- rownames(site_scores)  # Add sample names

species_scores <- as.data.frame(vegan::scores(rda_index, display = "species", choices = 1:3))
#species_scores <- as.data.frame(rda_index$CCA$v[, 1:3])
species_scores$Species <- rownames(species_scores)  # Add species names

# Extract biplot arrows (for explanatory variables)
biplot_arrows <- as.data.frame(vegan::scores(rda_index, display = "bp", choices = 1:3))
#biplot_arrows <- as.data.frame(rda_index$CCA$biplot[, 1:3])
biplot_arrows$variable <- rownames(biplot_arrows)

biplot_arrows <- biplot_arrows %>%
  mutate(variable = ifelse(variable == "meta_for_nmds$latitude", "latitude", variable),
         variable = ifelse(variable == "meta_for_nmds$depth", "depth", variable),
         variable = ifelse(variable == "as.factor(meta_for_nmds$collection_year)2023", "year", variable),
         variable = ifelse(variable == "meta_for_nmds$temp_C", "temperature", variable),
         variable = ifelse(variable == "meta_for_nmds$salinity", "salinity", variable))

## Plot CAP1 vs CAP2 - color sites by latitude
plot_lat <- ggplot() +
  # Plot site scores
  geom_point(data = site_scores, aes(x = CAP1, y = CAP2, color = meta_for_nmds$latitude), size = 3) +
  #geom_text_repel(data = site_scores, aes(x = CAP1, y = CAP2, label = Sample), size = 4) +
  scale_color_viridis(option = "cividis",  direction = -1) +
  
  # Plot species scores
  geom_point(data = species_scores, aes(x = CAP1, y = CAP2), color = "red", size = 2) +
  geom_text_repel(data = species_scores, aes(x = CAP1, y = CAP2, label = Species), color = "red", size = 2) +
  
  # Plot arrows
  geom_segment(data = biplot_arrows, aes(x = 0, y = 0, xend = CAP1, yend = CAP2), 
               arrow = arrow(length = unit(0.3, "cm")), color = "black") + # Plot biplot arrows
  geom_text(data = biplot_arrows, aes(x = CAP1 * 1.1, y = CAP2 * 1.1, label = variable), color = "black") + # Label arrows
  
  # Add labels and theme
  labs(title = "CAP Plot", 
       x = paste("CAP1 (", round(proportion_variance[1], 2), "% variance explained)", sep = ""), 
       y = paste("CAP2 (", round(proportion_variance[2], 2), "% variance explained)", sep = ""),
       color = "latitude") +
  theme_minimal() #+
  #theme(legend.position = "none")


## Plot CAP1 vs CAP2 - color sites by year
plot_year <- ggplot() +
  # Plot site scores
  geom_point(data = site_scores, aes(x = CAP1, y = CAP2, color = as.factor(meta_for_nmds$collection_year)), size = 3) +
  #geom_text_repel(data = site_scores, aes(x = CAP1, y = CAP2, label = Sample), size = 4) +
  
  # Plot species scores
  geom_point(data = species_scores, aes(x = CAP1, y = CAP2), color = "black", size = 2) +
  geom_text_repel(data = species_scores, aes(x = CAP1, y = CAP2, label = Species), color = "black", size = 2) +
  
  # Plot arrows
  geom_segment(data = biplot_arrows, aes(x = 0, y = 0, xend = CAP1, yend = CAP2), 
               arrow = arrow(length = unit(0.3, "cm")), color = "black") + # Plot biplot arrows
  geom_text(data = biplot_arrows, aes(x = CAP1 * 1.1, y = CAP2 * 1.1, label = variable), color = "black") + # Label arrows
  
  # Add labels and theme
  labs(title = "CAP Plot", 
       x = paste("CAP1 (", round(proportion_variance[1], 2), "% variance explained)", sep = ""), 
       y = paste("CAP2 (", round(proportion_variance[2], 2), "% variance explained)", sep = ""),  
       color = "year") +
  theme_minimal() #+
  #theme(legend.position = "none")


## Plot CAP1 vs CAP3 - color sites by depth
plot_depth <- ggplot() +
  # Plot site scores
  geom_point(data = site_scores, aes(x = CAP1, y = CAP3, color = meta_for_nmds$depth), size = 3) +
  #geom_text_repel(data = site_scores, aes(x = CAP1, y = CAP3, label = Sample), size = 4) +
  scale_colour_viridis(option = "C", direction = -1) + 
  
  # Plot species scores
  geom_point(data = species_scores, aes(x = CAP1, y = CAP3), color = "black", size = 2) +
  geom_text_repel(data = species_scores, aes(x = CAP1, y = CAP3, label = Species), color = "black", size = 2) +
  
  # Plot arrows
  geom_segment(data = biplot_arrows, aes(x = 0, y = 0, xend = CAP1, yend = CAP3), 
               arrow = arrow(length = unit(0.3, "cm")), color = "black") + # Plot biplot arrows
  geom_text(data = biplot_arrows, aes(x = CAP1 * 1.1, y = CAP3 * 1.1, label = variable), color = "black") + # Label arrows
  
  # Add labels and theme
  labs(title = "CAP Plot", 
       x = paste("CAP1 (", round(proportion_variance[1], 2), "% variance explained)", sep = ""), 
       y = paste("CAP3 (", round(proportion_variance[3], 2), "% variance explained)", sep = ""), 
       color = "depth") +
  theme_minimal() #+
  #theme(legend.position = "none")

## Plot CAP1 vs CAP3 - color sites by temp
plot_temp <- ggplot() +
  # Plot site scores
  geom_point(data = site_scores, aes(x = CAP1, y = CAP3, color = meta_for_nmds$temp_C), size = 3) +
  #geom_text_repel(data = site_scores, aes(x = CAP1, y = CAP3, label = Sample), size = 4) +
  scale_colour_viridis(option = "C", direction = -1) + 
  
  # Plot species scores
  geom_point(data = species_scores, aes(x = CAP1, y = CAP3), color = "black", size = 2) +
  geom_text_repel(data = species_scores, aes(x = CAP1, y = CAP3, label = Species), color = "black", size = 2) +
  
  # Plot arrows
  geom_segment(data = biplot_arrows, aes(x = 0, y = 0, xend = CAP1, yend = CAP3), 
               arrow = arrow(length = unit(0.3, "cm")), color = "black") + # Plot biplot arrows
  geom_text(data = biplot_arrows, aes(x = CAP1 * 1.1, y = CAP3 * 1.1, label = variable), color = "black") + # Label arrows
  
  # Add labels and theme
  labs(title = "CAP Plot", 
       x = paste("CAP1 (", round(proportion_variance[1], 2), "% variance explained)", sep = ""), 
       y = paste("CAP3 (", round(proportion_variance[3], 2), "% variance explained)", sep = ""), 
       color = "temperature") +
  theme_minimal() #+
  #theme(legend.position = "none")
```

```{r}
plot_lat
plot_year
plot_depth
```



```{r}
#ggsave(plot= plot_lat, "/home/kimberly.ledger/dbo_metabarcoding/outputs/latitude_fishonly.png", dpi = 300, width = 8, height = 6)
#ggsave(plot= plot_year, "/home/kimberly.ledger/dbo_metabarcoding/outputs/year_fishonly.png", dpi = 300, width = 8, height = 6)
ggsave(plot= plot_depth, "/home/kimberly.ledger/dbo_metabarcoding/outputs/depth_fishonly.png", dpi = 300, width = 8, height = 6)
```

now that i have some pretty plot, let's see the stats 

model summary 
```{r}
summary(rda_index) 
```


significance testing
```{r}
# Permutation test
anova_results <- anova(rda_index, permutations = how(nperm = 999))  # 999 permutations
anova_terms <- anova(rda_index, by = "terms", permutations = 999)
print(anova_results)
print(anova_terms)

# Get adjusted R² for the overall model
R2_adj <- RsquareAdj(rda_index)
cat("Adjusted R² for the overall model: ", R2_adj$adj.r.squared, "\n")
```

double check collinearity of explanatory variables - anything >10 should not be in the same model 
```{r}
# Variance Inflation Factors (VIF)
vif_values <- vif.cca(rda_index)
print(vif_values)
```

## test the same question but use GAM - it's a seperate model for each species 
```{r}
library(mgcv)


# Loop through each species column in your data frame
species_names <- colnames(index_df_wide)[-1]  # Exclude the first column (e.g., site or sample ID)

results <- list()  # To store the GAM models

for (species in species_names) {
  response_var <- index_df_wide[[species]]
  
  # Fit the GAM for each species
  gam_model <- gam(response_var ~ s(latitude) + s(depth) + s(temp_C) + as.factor(collection_year), 
                   data = meta_for_nmds)
  
  results[[species]] <- summary(gam_model)  # Store the summary of the model
}

# Access the results for each species
results$`Gadus chalcogrammus`
```


#identify species driving difference between lat, depth, year, etc. 

for categorial vars, i can use indval function from labdsv package 

```{r}
library(labdsv)

# Perform indicator species analysis
#indicator_year <- indval(index_df_wide[,-1], meta_for_nmds$collection_year, perm = 999)
indicator_year <- indval(binary_df_wide[,-1], meta_for_nmds$collection_year, perm = 999)
summary(indicator_year)

## there's a whole bunch more species in 2023 than in 2021
```

```{r}
# this only works for categorical variables, so let me make some dummy groups for this 
#quantile(meta_for_nmds$latitude, probs = seq(0, 1, length.out = 4))
binned_lat <- cut(meta_for_nmds$latitude, breaks = quantile(meta_for_nmds$latitude, probs = seq(0,1,0.25)), include.lowest = TRUE)

meta_for_nmds <- meta_for_nmds %>%
  mutate(latitude_group = ifelse(latitude <= 65, "low", NA),
         latitude_group = ifelse(latitude > 65 & latitude < 71, "middle", latitude_group),
         latitude_group = ifelse(latitude >= 71, "high", latitude_group))

#indicator_latitude<- indval(index_df_wide[,-1], meta_for_nmds$latitude_group, perm = 999)
#indicator_latitude<- indval(index_df_wide[,-1], binned_lat, perm = 999)

indicator_latitude<- indval(binary_df_wide[,-1], binned_lat, perm = 999)
indicator_latitude
summary(indicator_latitude)
```

using depth categories
```{r}
indicator_depth <- indval(index_df_wide[,-1], meta_for_nmds$depth_cat, perm = 999)
#indicator_depth <- indval(binary_df_wide[,-1], meta_for_nmds$depth_cat, perm = 999)
summary(indicator_depth)
```


taxa richness between 2021 and 2023 (fish only)
```{r}
taxon_table_with_year <- cbind(meta_for_nmds$collection_year, index_df_wide[,-1])
colnames(taxon_table_with_year)[1] <- "year"  # Rename the year column

# Exclude the "year" column and calculate species richness using rowSums
taxon_table_with_year <- taxon_table_with_year %>%
  mutate(richness = rowSums(select(., -year) > 0))

# Group by year and summarize richness
richness_by_year <- taxon_table_with_year %>%
  group_by(year) %>%
  summarise(mean_richness = mean(richness),
            total_richness = sum(richness),
            median_richness = median(richness),
            n_samples = n())  # Number of samples per year

richness_by_year
```

taxa richness (fish only) by year and latitude and depth
```{r}
taxon_table_with_year_and_lat <- cbind(meta_for_nmds$collection_year, meta_for_nmds$latitude, meta_for_nmds$depth, meta_for_nmds$temp_C, index_df_wide[,-1])
colnames(taxon_table_with_year_and_lat)[1] <- "year"  # Rename the year column
colnames(taxon_table_with_year_and_lat)[2] <- "latitude"
colnames(taxon_table_with_year_and_lat)[3] <- "depth"
colnames(taxon_table_with_year_and_lat)[4] <- "temp"

# Exclude the "year" column and calculate species richness using rowSums
taxon_table_with_year_and_lat <- taxon_table_with_year_and_lat %>%
  mutate(richness = rowSums(select(., -c(year,latitude,depth)) > 0))

ggplot(taxon_table_with_year_and_lat, aes(x = latitude, y = richness, color = as.factor(year))) +
  geom_point() +   # Scatter plot points
  geom_smooth(method = "loess", se = TRUE) +  # Smoothed line (loess curve)
  labs(x = "Latitude", y = "Taxa Richness", title = "Species Richness by Latitude", color = "latitude") +
  theme_minimal()                 # Apply a clean minimal theme

ggplot(taxon_table_with_year_and_lat, aes(x = temp, y = richness, color = latitude)) +
  geom_point() +   # Scatter plot points
  scale_colour_viridis(option = "C", direction = -1) +
  geom_smooth(method = "loess", se = TRUE) +  # Smoothed line (loess curve)
  labs(x = "Temperature", y = "Taxa Richness", title = "Species Richness by Temperature", color = "Latitude") +
  theme_minimal()                 # Apply a clean minimal theme

```

run a GAM modeling species richness by latitude, year and depth 
```{r}
# Fit a GAM on species richness
gam_model_richness <- gam(richness ~ s(latitude) + s(depth) + s(temp) + as.factor(year), 
                           data = taxon_table_with_year_and_lat)

summary(gam_model_richness)
plot(gam_model_richness)
```

