---
title: "dbo gadids"
output: html_document
date: "2024-10-16"
---

just looking at the gadid detection in the dbo mifish metabarcoding dataset 

some thoughts: since the metabarcoding data targeted all fish, not just gadids, that means the number of reads or proportion of gadid reads per sample is influenced by the non-gadid species present. therefore, i believe the appropriate approach would be to use gadid reads as binary data. and i will try out some proportion of positive technical replicate response variable for bottles/sites to act as a rough proxy for the amount of DNA recovered. 

libraries
```{r}
library(tidyverse)
library(ggplot2)
rename <- dplyr::rename
#library(viridis)
#library(ggrepel)
```

the data 
```{r}
taxon_table <- read.csv("/home/kimberly.ledger/dbo_metabarcoding/outputs/decontaminated_taxon_table_filtered.csv") %>%
  select(!X)  %>%
  filter(sample_type == "sample")

metadata <- read.csv("/home/kimberly.ledger/dbo_metabarcoding/data/NBS_SBS_DBO_metadata.csv") %>%
  filter(project == "DBO")

metadata_mini <- metadata %>%
  select(extraction_ID, alternative_ID, sample_type, project, collection_year, location1, location2, depth, longitude, latitude) %>%
  unite("project_year", project, collection_year, sep = "_", remove = F) %>%
  mutate(extraction_ID = ifelse(project_year == "DBO_2021", alternative_ID, extraction_ID)) %>% # for DBO2021 samples, replace extraction_ID with alternative_ID
  select(!alternative_ID)
```

get additional cruise metadata - this is not a final version for 2023 samples 
```{r}
cruise_meta <- read.csv("/home/kimberly.ledger/dbo_metabarcoding/data/SikuliaqDBO_2021_2023_cruisedata_20241015.csv") %>%
  mutate(temp_C = (temperature_ch1 + temperature_ch2)/2) %>%
  mutate(salinity = (salinity_ch1 + salinity_ch2)/2)
```

```{r}
metadata_join <- metadata_mini %>%
  left_join(cruise_meta) %>%
  mutate(temp_C = ifelse(extraction_ID == "e04213", -0.89975, temp_C),   ## filling in missing data by taking the average on the station before and after at same depth 
         temp_C = ifelse(extraction_ID == "e04214", 2.77975, temp_C),
         salinity = ifelse(extraction_ID == "e04213", 32.45635, salinity),
         salinity = ifelse(extraction_ID == "e04214", 30.39102, salinity),
         Collection_Time_local = ifelse(extraction_ID == "E1090.SKQ2021", '10:31', Collection_Time_local),
         Collection_Time_local = ifelse(extraction_ID == "E1088.SKQ2021", '10:31', Collection_Time_local),
         Collection_Time_local = ifelse(extraction_ID == "E1089.SKQ2021", '10:31', Collection_Time_local))
```

filter taxon table to keep just the gadids 
```{r}
gadids <- taxon_table %>%
  filter(family == "Gadidae")

gadids %>%
  group_by(collection_year, taxon) %>%
  summarise(n_reads = sum(tot_reads)) %>%
  group_by(collection_year) %>%
  mutate(gadidae_reads = sum(n_reads),
         prop_reads = n_reads/gadidae_reads) %>%
  select(collection_year, taxon, prop_reads) %>%
  pivot_wider(names_from = collection_year, values_from = prop_reads)
  
```

okay so gadidae reads make us <1% in each year and gadus made up 4-9% of reads 
mostly pollock (80% and 50%, respectively) 
way higher proportion of reads for arctic cod in 2023 than 2021 
pretty low proportions of pcod (~1-2%)

anyway, ~90% of reads go to species level so for now i'll move forward with just the reads that get down to species 

the dataframe i'm working with right now, is already the sum of the three technical replicates for 2023 data - would need to go back a few steps if i want to incorporate variability in tech replicates for 2023 samples (which might be worth it)

pivot gadid table 
```{r}
gadids_wide <- gadids[,c(1,6,13)] %>%
  pivot_wider(names_from = taxon, values_from = tot_reads)
```

join the gadid table to metadata to fill in extractions with no gadid reads 

```{r}
gadids_w_meta <- metadata_join %>%
  filter(sample_type == "sample") %>%
  left_join(gadids_wide) %>%
  replace_na(list('Arctogadus glacialis' = 0, 'Boreogadus saida' = 0, 'Eleginus gracilis' = 0, 'Gadidae' = 0, 'Gadus' = 0, 'Gadus chalcogrammus' = 0, 'Gadus macrocephalus' = 0))
```

pull out just the binary data, now that all rows/samples are there 
```{r}
extraction_ids <- gadids_w_meta$extraction_ID

tax_df <- gadids_w_meta[,c(26:32)] %>%
  mutate(across(where(is.numeric), ~ ifelse(. > 0, 1, .)))
```

for each station and year (i.e. combine bottles from multiple depths), count the total number of bottles and number of bottles with detections for each species 

```{r}
gadids_w_meta %>%
  group_by(collection_year, location1, `Arctogadus glacialis`, , `, `Gadus chalcogrammus`, `Gadus macrocephalus`) %>%
  summarize(n_bottles = n(),
            AG = sum())

pres <- gadids_w_meta %>%
  group_by(collection_year, location1) %>%
  summarize(
    n_bottles = n(),                            # Total number of rows
    n_AG = sum(`Arctogadus glacialis` > 0, na.rm = TRUE),   # Number of rows where AC > 0
    n_BS = sum(`Boreogadus saida` > 0, na.rm = TRUE), # Number of rows where BS > 0
    n_EG = sum(`Eleginus gracilis` > 0, na.rm = TRUE), # Number of rows where EG > 0
    n_GC = sum(`Gadus chalcogrammus` > 0, na.rm = TRUE), # Number of rows where GC > 0
    n_GM = sum(`Gadus macrocephalus` > 0, na.rm = TRUE),
    prop_AG = n_AG/n_bottles, 
    prop_BS = n_BS/n_bottles,
    prop_EG = n_EG/n_bottles,
    prop_GC = n_GC/n_bottles,
    prop_GM = n_GM/n_bottles)
```

get the year/location metadata 
```{r}
df <- gadids_w_meta %>%
  group_by(collection_year, location1, location2, longitude, latitude) %>%
  summarise(mean_temp = mean(temp_C, na.rm = T),
            mean_salinity = mean(salinity, na.rm = T)) %>%
  left_join(pres)
```


i'm not sure the mean temp is the appropriate thing to use in these models... but for now. 
```{r}
#Sys.setenv(PATH = paste("/opt/anaconda/bin", Sys.getenv("PATH"), sep = ":"))
#Sys.which("cmake")
#install.packages("glmmTMB")

library(glmmTMB)

arctic_c <- glmmTMB(cbind(n_BS, n_bottles - n_BS) ~ latitude + as.factor(collection_year), family = binomial(link = "logit"), 
                 data = df)

summary(arctic_c)

pollock <- glmmTMB(cbind(n_GC, n_bottles - n_GC) ~ latitude  + as.factor(collection_year), family = binomial(link = "logit"), 
                 data = df)

summary(pollock)

pcod <- glmmTMB(cbind(n_GM, n_bottles - n_GM) ~ latitude + mean_temp, family = binomial(link = "logit"), 
                 data = df)

summary(pcod)
```


okay, how about a visualization 

load libraries
```{r}
library(rnaturalearth)
library(rnaturalearthdata)
world <- ne_countries(scale = "medium", returnclass = "sf")
```


plot 
```{r}
min_lat <- min(df$latitude, na.rm = T)
max_lat <- max(df$latitude, na.rm = T)

min_lon <- min(df$longitude, na.rm = T)
max_lon <- max(df$longitude, na.rm = T)

df$collection_year <- as.factor(df$collection_year)

arctic_cod_plot <- ggplot(data = world) +
    geom_sf() +
    geom_point(data = df, aes(x = longitude, y = latitude, color= prop_BS)) + 
  coord_sf(xlim = c(min_lon-2, max_lon+2), ylim = c(min_lat-1, max_lat+1), expand = FALSE) +
  theme_bw() +xlab("Longitude") +ylab("Latitude") +
  facet_grid(~collection_year) + 
  #labs(title = df$taxon) +
  theme(axis.text.x = element_text(angle = 90))

arctic_cod_plot

pollock_plot <- ggplot(data = world) +
    geom_sf() +
    geom_point(data = df, aes(x = longitude, y = latitude, color= prop_GC)) + 
  coord_sf(xlim = c(min_lon-2, max_lon+2), ylim = c(min_lat-1, max_lat+1), expand = FALSE) +
  theme_bw() +xlab("Longitude") +ylab("Latitude") +
  facet_grid(~collection_year) + 
  #labs(title = df$taxon) +
  theme(axis.text.x = element_text(angle = 90))

pollock_plot

temp_plot <- ggplot(data = world) +
    geom_sf() +
    geom_point(data = df, aes(x = longitude, y = latitude, color= mean_temp)) + 
  coord_sf(xlim = c(min_lon-2, max_lon+2), ylim = c(min_lat-1, max_lat+1), expand = FALSE) +
  theme_bw() +xlab("Longitude") +ylab("Latitude") +
  facet_grid(~collection_year) + 
  #labs(title = df$taxon) +
  theme(axis.text.x = element_text(angle = 90))

temp_plot
```

try working with afsc maps

```{r}
#devtools::install_github("afsc-gap-products/akgfmaps", build_vignettes = TRUE)

#library(akgfmaps)
#browseVignettes('akgfmaps')
```


