---
title: "dbo gadids"
output: html_document
date: "2024-10-18"
---

just looking at the gadid detection in the dbo mifish metabarcoding dataset 

some thoughts: even though read proportions (or maybe an eDNA index) might not be the way to go here, let's try it out for comparison with the proportion of gadids in catch. 

and going to work with data version where 2023 pcr reps have been combined 

libraries
```{r}
library(tidyverse)
library(ggplot2)
rename <- dplyr::rename
```

the data 
```{r}
taxon_table <- read.csv("/home/kimberly.ledger/dbo_metabarcoding/outputs/decontaminated_taxon_table_filtered.csv") %>%
  select(!X)  %>%
  filter(sample_type == "sample")

metadata <- read.csv("/home/kimberly.ledger/dbo_metabarcoding/data/NBS_SBS_DBO_metadata.csv") %>%
  filter(project == "DBO")

metadata_mini <- metadata %>%
  select(extraction_ID, alternative_ID, sample_type, project, collection_year, location1, location2, depth, longitude, latitude) %>%
  unite("project_year", project, collection_year, sep = "_", remove = F) %>%
  mutate(extraction_ID = ifelse(project_year == "DBO_2021", alternative_ID, extraction_ID)) %>% # for DBO2021 samples, replace extraction_ID with alternative_ID
  select(!alternative_ID)
```

get additional cruise metadata - this is not a final version for 2023 samples 
```{r}
cruise_meta <- read.csv("/home/kimberly.ledger/dbo_metabarcoding/data/SikuliaqDBO_2021_2023_cruisedata_20241015.csv") %>%
  mutate(temp_C = (temperature_ch1 + temperature_ch2)/2) %>%
  mutate(salinity = (salinity_ch1 + salinity_ch2)/2)
```

```{r}
metadata_join <- metadata_mini %>%
  left_join(cruise_meta) %>%
  mutate(temp_C = ifelse(extraction_ID == "e04213", -0.89975, temp_C),   ## filling in missing data by taking the average on the station before and after at same depth 
         temp_C = ifelse(extraction_ID == "e04214", 2.77975, temp_C),
         salinity = ifelse(extraction_ID == "e04213", 32.45635, salinity),
         salinity = ifelse(extraction_ID == "e04214", 30.39102, salinity),
         Collection_Time_local = ifelse(extraction_ID == "E1090.SKQ2021", '10:31', Collection_Time_local),
         Collection_Time_local = ifelse(extraction_ID == "E1088.SKQ2021", '10:31', Collection_Time_local),
         Collection_Time_local = ifelse(extraction_ID == "E1089.SKQ2021", '10:31', Collection_Time_local))
```

filter taxon table to keep just the gadids 
```{r}
gadids <- taxon_table %>%
  filter(family == "Gadidae")

gadids %>%
  group_by(collection_year, taxon) %>%
  summarise(n_reads = sum(tot_reads)) %>%
  group_by(collection_year) %>%
  mutate(gadidae_reads = sum(n_reads),
         prop_reads = n_reads/gadidae_reads) %>%
  select(collection_year, taxon, prop_reads) %>%
  pivot_wider(names_from = collection_year, values_from = prop_reads)
```

okay so gadidae reads make us <1% in each year and gadus made up 4-9% of reads 
mostly pollock (80% and 50%, respectively) 
much higher proportion of reads for arctic cod in 2023 than 2021 
pretty low proportions of pcod (~1-2%) overall 

anyway, ~90% of reads go to species level so for now i'll move forward with just the reads that get down to species 

pivot gadid table 
```{r}
gadids_wide <- gadids[,c(1,6,13)] %>%
  pivot_wider(names_from = taxon, values_from = tot_reads)
```

join the gadid table to metadata to fill in extractions with no gadid reads 
```{r}
gadids_w_meta <- metadata_join %>%
  filter(sample_type == "sample") %>%
  left_join(gadids_wide) %>%
  replace_na(list('Arctogadus glacialis' = 0, 'Boreogadus saida' = 0, 'Eleginus gracilis' = 0, 'Gadidae' = 0, 'Gadus' = 0, 'Gadus chalcogrammus' = 0, 'Gadus macrocephalus' = 0))
```

for each station and year (i.e. combine bottles from multiple depths), summarize the data in the following ways: 
- count the total number of bottles and number of bottles with detections for each species 

```{r}
gadid_mb <- gadids_w_meta %>% 
   mutate(lat_bin = ifelse(latitude <= 64, "<64", NA),
         lat_bin = ifelse(latitude > 64 & latitude < 66, "64-66", lat_bin), 
         lat_bin = ifelse(latitude > 66 & latitude < 68, "66-68", lat_bin),
         lat_bin = ifelse(latitude > 68 & latitude < 70, "68-70", lat_bin),
         lat_bin = ifelse(latitude > 70 & latitude < 72, "70-72", lat_bin),
         lat_bin = ifelse(latitude >= 72, ">72", lat_bin)) %>%
  mutate(AC_pres = ifelse(`Boreogadus saida` > 0, 1, 0),
         WP_pres = ifelse(`Gadus chalcogrammus` > 0, 1, 0),
         PC_pres = ifelse(`Gadus macrocephalus` > 0, 1, 0),
         SC_pres = ifelse(`Eleginus gracilis` > 0, 1, 0),
         PolC_pres = ifelse(`Arctogadus glacialis` > 0, 1, 0)) %>%
  group_by(collection_year, lat_bin) %>%
  summarise(unique_bottles = n(),
            AC = sum(AC_pres),
            WP = sum(WP_pres),
            PC = sum(PC_pres),
            SC = sum(SC_pres),
            PolC = sum(PolC_pres),
            prop_AC = AC/unique_bottles,
            prop_WP = WP/unique_bottles,
            prop_PC = PC/unique_bottles,
            prop_SC = SC/unique_bottles,
            prop_PolC = PolC/unique_bottles)

my_order <- c("<64", "64-66", "66-68", "68-70", "70-72", ">72")
gadid_mb$lat_bin <- factor(gadid_mb$lat_bin, levels = my_order)
  
mb_prop_w_pos <- gadid_mb %>%
  select(collection_year, lat_bin, prop_AC, prop_WP, prop_PC, prop_SC, prop_PolC) %>%
  pivot_longer(cols = c(3:7), values_to = "prop", names_to = "taxon") %>%
  mutate(CommonName = ifelse(taxon == "prop_AC", "Arctic cod", NA),
         CommonName = ifelse(taxon == "prop_WP", "Walleye pollock", CommonName),
         CommonName = ifelse(taxon == "prop_PC", "Pacific cod", CommonName),
         CommonName = ifelse(taxon == "prop_SC", "Saffron cod", CommonName),
         CommonName = ifelse(taxon == "prop_PolC", "Polar cod", CommonName))
```


```{r}
mb_prop_w_pos %>%
  ggplot(aes(x=CommonName, y=lat_bin, fill = prop)) +
  geom_tile() + 
  scale_fill_gradient(low = "lightblue", high = "darkblue") +  # Adjust color scale as needed
  facet_grid(~collection_year) + 
  labs(title = "Heatmap of Species Proportion by Latitude",
       x = "Species",
       y = "Latitude",
       fill = "Proportion of Total Detection in MB") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability
```

how many bottle of water where collected in each year at each lat bin? 
```{r}
gadid_mb %>%
  select(collection_year, lat_bin, unique_bottles)
```




get the year/location metadata 
```{r}
df <- gadids_w_meta %>%
  group_by(collection_year, location1, location2, longitude, latitude) %>%
  summarise(mean_temp = mean(temp_C, na.rm = T),
            mean_salinity = mean(salinity, na.rm = T)) %>%
  left_join(pres)
```


i'm not sure the mean temp is the appropriate thing to use in these models... but for now. 
```{r}
#Sys.setenv(PATH = paste("/opt/anaconda/bin", Sys.getenv("PATH"), sep = ":"))
#Sys.which("cmake")
#install.packages("glmmTMB")

library(glmmTMB)

arctic_c <- glmmTMB(cbind(n_BS, n_bottles - n_BS) ~ latitude + as.factor(collection_year), family = binomial(link = "logit"), 
                 data = df)

summary(arctic_c)

pollock <- glmmTMB(cbind(n_GC, n_bottles - n_GC) ~ latitude  + as.factor(collection_year), family = binomial(link = "logit"), 
                 data = df)

summary(pollock)

pcod <- glmmTMB(cbind(n_GM, n_bottles - n_GM) ~ latitude + mean_temp, family = binomial(link = "logit"), 
                 data = df)

summary(pcod)
```


okay, how about a visualization 

load libraries
```{r}
library(rnaturalearth)
library(rnaturalearthdata)
world <- ne_countries(scale = "medium", returnclass = "sf")
```


plot 
```{r}
min_lat <- min(df$latitude, na.rm = T)
max_lat <- max(df$latitude, na.rm = T)

min_lon <- min(df$longitude, na.rm = T)
max_lon <- max(df$longitude, na.rm = T)

df$collection_year <- as.factor(df$collection_year)

arctic_cod_plot <- ggplot(data = world) +
    geom_sf() +
    geom_point(data = df, aes(x = longitude, y = latitude, color= prop_BS)) + 
  coord_sf(xlim = c(min_lon-2, max_lon+2), ylim = c(min_lat-1, max_lat+1), expand = FALSE) +
  theme_bw() +xlab("Longitude") +ylab("Latitude") +
  facet_grid(~collection_year) + 
  #labs(title = df$taxon) +
  theme(axis.text.x = element_text(angle = 90))

arctic_cod_plot

pollock_plot <- ggplot(data = world) +
    geom_sf() +
    geom_point(data = df, aes(x = longitude, y = latitude, color= prop_GC)) + 
  coord_sf(xlim = c(min_lon-2, max_lon+2), ylim = c(min_lat-1, max_lat+1), expand = FALSE) +
  theme_bw() +xlab("Longitude") +ylab("Latitude") +
  facet_grid(~collection_year) + 
  #labs(title = df$taxon) +
  theme(axis.text.x = element_text(angle = 90))

pollock_plot

saffron_plot <- ggplot(data = world) +
    geom_sf() +
    geom_point(data = df, aes(x = longitude, y = latitude, color= prop_EG)) + 
  coord_sf(xlim = c(min_lon-2, max_lon+2), ylim = c(min_lat-1, max_lat+1), expand = FALSE) +
  theme_bw() +xlab("Longitude") +ylab("Latitude") +
  facet_grid(~collection_year) + 
  #labs(title = df$taxon) +
  theme(axis.text.x = element_text(angle = 90))

saffron_plot

pcod_plot <- ggplot(data = world) +
    geom_sf() +
    geom_point(data = df, aes(x = longitude, y = latitude, color= prop_GM)) + 
  coord_sf(xlim = c(min_lon-2, max_lon+2), ylim = c(min_lat-1, max_lat+1), expand = FALSE) +
  theme_bw() +xlab("Longitude") +ylab("Latitude") +
  facet_grid(~collection_year) + 
  #labs(title = df$taxon) +
  theme(axis.text.x = element_text(angle = 90))

pcod_plot

temp_plot <- ggplot(data = world) +
    geom_sf() +
    geom_point(data = df, aes(x = longitude, y = latitude, color= mean_temp)) + 
  coord_sf(xlim = c(min_lon-2, max_lon+2), ylim = c(min_lat-1, max_lat+1), expand = FALSE) +
  theme_bw() +xlab("Longitude") +ylab("Latitude") +
  facet_grid(~collection_year) + 
  #labs(title = df$taxon) +
  theme(axis.text.x = element_text(angle = 90))

temp_plot
```

plot presence/absence
```{r}
min_lat <- min(df$latitude, na.rm = T)
max_lat <- max(df$latitude, na.rm = T)

min_lon <- min(df$longitude, na.rm = T)
max_lon <- max(df$longitude, na.rm = T)

df$collection_year <- as.factor(df$collection_year)

arctic_cod_plot <- ggplot(data = world) +
    geom_sf() +
    geom_point(data = df, aes(x = longitude, y = latitude, color= as.factor(BS))) + 
  coord_sf(xlim = c(min_lon-2, max_lon+2), ylim = c(min_lat-1, max_lat+1), expand = FALSE) +
  theme_bw() +xlab("Longitude") +ylab("Latitude") +
  facet_grid(~collection_year) + 
  #labs(title = df$taxon) +
  theme(axis.text.x = element_text(angle = 90))

arctic_cod_plot

pollock_plot <- ggplot(data = world) +
    geom_sf() +
    geom_point(data = df, aes(x = longitude, y = latitude, color= as.factor(GC))) + 
  coord_sf(xlim = c(min_lon-2, max_lon+2), ylim = c(min_lat-1, max_lat+1), expand = FALSE) +
  theme_bw() +xlab("Longitude") +ylab("Latitude") +
  facet_grid(~collection_year) + 
  #labs(title = df$taxon) +
  theme(axis.text.x = element_text(angle = 90))

pollock_plot


pcod_plot <- ggplot(data = world) +
    geom_sf() +
    geom_point(data = df, aes(x = longitude, y = latitude, color= as.factor(GM))) + 
  coord_sf(xlim = c(min_lon-2, max_lon+2), ylim = c(min_lat-1, max_lat+1), expand = FALSE) +
  theme_bw() +xlab("Longitude") +ylab("Latitude") +
  facet_grid(~collection_year) + 
  #labs(title = df$taxon) +
  theme(axis.text.x = element_text(angle = 90))

pcod_plot

```



try working with afsc maps

```{r}
#devtools::install_github("afsc-gap-products/akgfmaps", build_vignettes = TRUE)

#library(akgfmaps)
#browseVignettes('akgfmaps')
```


