---
title: "chukchi catch"
output: html_document
date: "2024-10-22"
---

updated 10/29/24 - using new BASIS data export from Andrew Diamond 

```{r}
library(tidyverse)
library(taxize)
```

read in BASIS catch data from >60N in years of Integrated Arctic Surveys
```{r}
chukchi_catch <- read.csv("/home/kimberly.ledger/dbo_metabarcoding/data/Kim_Arctic_10.28.24.csv") %>%
  filter(!StationID %in% c("20171602001", "20171602002", "20171602003"))  # remove three NBS stations below 60N that were included in data export

# what unique cruises are in this data and how many station did each sample? 
chukchi_catch %>%
  separate(StationID, into = c("year_vessel_cruise", "station"), sep = 8, remove = F) %>%
  group_by(year_vessel_cruise) %>%
  summarize(n_stations = n_distinct(station))
```

get full taxonomic info for the taxa caught
```{r}
# taxa list for this catch data 
taxa_list <- unique(chukchi_catch$ScientificName)

# look up the tsn codes of basis surveys  
tsn_codes <- httr::content(
    httr::GET('https://apex.psmfc.org/akfin/data_marts/akmp/basis_tsn'),
    type = "application/json") %>%
  bind_rows() %>%
  rename_with(tolower) %>%
  dplyr::rename(speciestsn = tsn) %>%
  select(speciestsn, commonname, scientificname) %>%
  filter(scientificname %in% taxa_list)

full_classification <- classification(tsn_codes$speciestsn, db = "itis")

# Remove any NULL or empty results from the list
full_classification_clean <- full_classification  %>%
  keep(~ !is.null(.) && is.data.frame(.) && nrow(.) > 0)

# Convert the list of taxonomies into a dataframe
taxonomy_df <- full_classification_clean %>%
  bind_rows(.id = "speciestsn") %>%
  select(!id) %>%
  pivot_wider(names_from = rank, values_from = name) %>%
  select(speciestsn, class, order, family, genus, species) %>%
  filter(speciestsn != 171582) #remove duplicate entry

# join to my taxa list 
taxa_df <- tsn_codes %>%
  left_join(taxonomy_df, by = "speciestsn")

# i am going to filter by class == "Teleostei" since that's what the MiFish primer's target, do any of the entries with class == NA need to stay? 
taxa_na <- taxa_df %>%
  filter(is.na(class))
taxa_na
```

okay, so sturgeon poacher's scientific name is misspelled 
```{r}
missing_taxa <- data.frame(matrix(ncol = ncol(taxa_na), nrow = 6))
colnames(missing_taxa) <- colnames(taxa_na)

missing_taxa$speciestsn <- c("167481", "171582", "171584", "171587", "167281", "167373")
missing_taxa$commonname <- c("Sturgeon Poacher", "Slender eelblenny", "Daubed shanny", "Stout eelblenny", "Butterfly sculpin", "Ribbed Sculpin")
missing_taxa$scientificname <- c("Podothecus acipenserinus", "Lumpenus fabricii", "Lumpenus maculatus", "Lumpenus medius", "Hemilepidotus papilio", "Triglops pingeli")
missing_taxa$class <- c("Teleostei", "Teleostei", "Teleostei", "Teleostei", "Teleostei", "Teleostei") 
missing_taxa$order <- c("Perciformes", "Perciformes", "Perciformes", "Perciformes", "Perciformes", "Perciformes")
missing_taxa$family <- c("Agonidae", "Lumpenidae", "Lumpenidae", "Lumpenidae", "Cottidae", "Cottidae")
missing_taxa$genus <- c("Podothecus", "Lumpenus", "Lumpenus", "Lumpenus", "Hemilepidotus", "Triglops")
missing_taxa$species <- c("Podothecus acipenserinus", "Lumpenus fabricii", "Lumpenus maculatus", "Lumpenus medius", "Hemilepidotus papilio", "Triglops pingeli")

# Update `taxa_df` with values from `missing_taxa` for matching `speciestsn`
updated_taxa_df <- taxa_df %>%
  rows_update(missing_taxa, by = "speciestsn")

updated_taxa_df %>%
  filter(is.na(class))
```

join full taxonomic information to the catch data and filter to just retain the fish 
```{r}
fish_catch <- chukchi_catch %>%
  dplyr::rename(commonname = "CommonName") %>%
  dplyr::rename(scientificname = "ScientificName") %>%
  left_join(updated_taxa_df) %>%
  filter(class == "Teleostei")
```

how many taxa are identified to the species level, and if not, what are the IDs? 
```{r}
fish_catch %>%
  filter(scientificname == species) %>%
  select(commonname, scientificname) %>%
  unique()

fish_catch %>%
  filter(scientificname != species | is.na(species) | is.na(genus) | is.na(family) | is.na(order)) %>%
  select(commonname, scientificname, species) %>%
  unique()
```

okay, let's make a taxon column that's analogous to the eDNA mb data that is the lowest taxonomic resolution ID 

```{r}
fish_catch <- fish_catch %>%
  mutate(taxon = ifelse(scientificname == species, species, NA),
         taxon = ifelse(scientificname == "Osmerus mordax dentex", species, taxon),
         taxon = ifelse(scientificname == "Clupea pallasi", species, taxon),
         taxon = ifelse(scientificname == "Lumpenus maculatus", species, taxon),
         taxon = ifelse(scientificname == "Myoxocephalus scorpiodes", species, taxon),
         taxon = ifelse(scientificname == "Eumesogrammus preacisus", species, taxon))

fish_catch <- fish_catch %>%
  mutate(taxon = ifelse(is.na(taxon) & !is.na(genus), genus, taxon)) %>%
  mutate(taxon = ifelse(is.na(taxon) & !is.na(family), family, taxon)) %>%
  filter(!is.na(taxon))  #this removes just a few unidentified fish and flatfish 

fish_catch %>%
  select(commonname, scientificname, order, family, genus, species, taxon) %>%
  unique()  
```

hooray! taxon ids are looking good, but let's verify that all the species are known to occur in the North Pacific and/or Pacific Arctic. i'm not going to worry about taxa that are above species level... 

```{r}
library(rfishbase)

#one-time download of all fishbase tables... this takes a bit 
fb_import()

spp_list <- unique(fish_catch$species)

#first, validate species names using rfishbase synonyms
spp_df <- synonyms(spp_list)

#first use synonym column 

#get distribution info 
spp_distribution <- distribution(spp_df$Species) %>%
  select(Species, FAO) %>%
  unique()

#add column to designate if we will consider a species as "in range"- for this study, this will be NE Pacific and Arctic Ocean 
spp_distribution_range <- spp_distribution %>%
  mutate(in_range = ifelse(is.na(FAO), NA, "no"),
         in_range = ifelse(FAO == "Pacific, Northeast", "yes", in_range),
         in_range = ifelse(FAO == "Arctic Ocean", "yes", in_range))

#keep just a list of spp names and yes/no/NA for "in range"  - this way we can keep track of what spp didn't have any reference information in fishbase to determine range 
spp_range <- spp_distribution_range %>%
  select(Species, in_range) %>%
  unique() %>%
  filter(!is.na(Species))

# Find species with "no" in_range but no "yes" for the same species
no_without_yes <- spp_range %>%
  group_by(Species) %>%
  filter(all(in_range == "no")) %>%
  distinct(Species)

# check that none of these species known as Atlantic Arctic Species 
atlantic_arctic_spp <- read.csv("/home/kimberly.ledger/BeringSea_Arctic_eDNA/data/AtlanticArctic_SpeciesList.csv")

spp_df %>%
  filter(Species %in% atlantic_arctic_spp$Name)
```

cool, cool taxonomy looks good to go 

let's make a quick summary table of the fish catch 
```{r}
fish_catch %>%
  separate(StationID, into = c("year_vessel_cruise", "station"), sep = 8, remove = F) %>%
  group_by(year_vessel_cruise, taxon) %>%
  summarize(sum_TotalCatchNum = sum(TotalCatchNum),
            sum_TotalCatchWt = sum(TotalCatchWt))
```

alright, now since i know that the field IDs of 2017 and 2017 gadids were not great, I will reassign age-0 gadids based on the genetic ID proportions 

```{r}
gadid_age0 <- fish_catch %>%
  filter(SampleYear %in% c(2017, 2019)) %>%
  filter(family == "Gadidae") %>%
  filter(LHSCode == "A0") %>%                 #### come back and check/add U after Andrew looks at the data
  filter(TowType == "Surface")
```

read in gadid catch datafile that i reformatted to better work with in R 
```{r}
genetic_ids_2017 <- read.csv("/home/kimberly.ledger/dbo_metabarcoding/data/gadidsamples2010to2019.csv") %>%
  #filter(collection_year %in% c(2017, 2019)) #%>%
  filter(collection_year %in% c(2017)) %>%
  dplyr::rename(EQ_Latitude = StartLatitude_N) %>%
  dplyr::rename(EQ_Longitude = StartLongitude_W)
```

try to identify the trawl station IDs associated with the catch dataset using the lat/long
```{r}
station_locations_2017 <- fish_catch %>%
  #filter(SampleYear %in% c(2017, 2019)) %>%
  filter(SampleYear %in% c(2017)) %>%
  select(StationID, TowType, EQ_Latitude, EQ_Longitude) %>%
  unique()

genetic_ids_2017 <- genetic_ids_2017 %>%
  left_join(station_locations_2017)

## station where lat/long did not match 
no_match_2017 <- genetic_ids_2017 %>%
  filter(is.na(StationID)) %>%
  select(EQ_Latitude, EQ_Longitude) %>%
  unique()

## maybe a fuzzyjoin will work..
library(fuzzyjoin)

genetic_ids_2017_no_match <- genetic_ids_2017 %>%
  filter(is.na(StationID)) %>%
  select(AlternateID:EQ_Longitude)
  
# Join with a tolerance of ±0.1 for latitude and longitude
matched_df_2017 <- fuzzy_left_join(
  genetic_ids_2017_no_match, station_locations_2017,
  by = c("EQ_Latitude" = "EQ_Latitude", "EQ_Longitude" = "EQ_Longitude"),
  match_fun = list(
    function(x, y) abs(x - y) <= 0.3,
    function(x, y) abs(x - y) <= 0.3
  )
)

matched_df_2017 %>%
  filter(is.na(StationID)) %>%
  select(EQ_Latitude.x, EQ_Longitude.x) %>%
  unique()


#write.csv(matched_df, "/home/kimberly.ledger/dbo_metabarcoding/outputs/genetic_id_w_stations.csv")
```





```{r}
fish_catch %>%
  filter(order == "Gadiformes") %>%
  select(commonname) %>%
  unique()
```


pivot wider and remove 2010 samples and stations <60N
```{r}
fish_wide <- fish_catch %>%
  filter(sampleyear != 2010) %>%
  filter(!is.na(scientificname)) %>%
  filter(!is.na(totalcatchnum)) %>%
  filter(eq_latitude >= 60) %>%
  select(stationid, totalcatchnum, scientificname) %>%
  pivot_wider(names_from = scientificname, values_from = totalcatchnum, values_fn = sum, values_fill = 0)
```

station metadata of interest - will need to see if i can track down temp, salinity, etc..
need to ask someone about sample method code 
```{r}
chuk_meta <- chukchi_catch %>%
  filter(sampleyear != 2010) %>%
  filter(eq_latitude >= 60) %>%
  select(stationid, sampleyear, vesselcode.x, stationnumber.x, hauldate, eq_latitude, eq_longitude) %>%
  unique()
```

leftjoin the catch to the metadata to fill in trawls with nothing
```{r}
fish_wide_all_stations <- chuk_meta %>%
  left_join(fish_wide, by = "stationid")
```

okay, for multivariate analyses i can't have any empty rows (stations with no fish) so i'll need to remove those 
```{r}
fish_wide_filtered <- fish_wide_all_stations %>%
  filter(rowSums(select(., 8:ncol(.)), na.rm = TRUE) != 0)            

fish_totalcatchnum <- fish_wide_filtered[,-c(1:7)] 

fish_binary <- fish_wide_filtered[,-c(1:7)] %>%
   mutate(across(everything(), ~ ifelse(is.na(.), 0, ifelse(. > 0, 1, 0))))
  
fish_meta <- fish_wide_filtered[,c(1:7)]  

binary_dist<- vegdist(fish_binary, method = "jaccard")

# Run PERMANOVA with latitude as the predictor
binary_lat <- adonis2(binary_dist ~ fish_meta$eq_latitude, permutations = 999)
binary_lat
binary_year <- adonis2(binary_dist ~ as.factor(fish_meta$sampleyear), permutations = 999)
binary_year
```

check dispersion for sampling year
```{r}
dispersion <- betadisper(binary_dist, as.factor(fish_meta$sampleyear))

# Step 2: Perform a permutation test for multivariate homogeneity of group dispersions (PERMDISP)
dispersion_test <- permutest(dispersion, permutations = 999)

# Step 3: View the results of the test
print(dispersion_test)

# Step 4: Optionally, visualize the distances to group centroids
plot(dispersion)
boxplot(dispersion, main = "Group Dispersion by Year")
```


# Fit the dbRDA model - using capscale() - constrained analysis of principal coordinates
```{r}
rda_index <- capscale(fish_binary ~ fish_meta$eq_latitude + as.factor(fish_meta$sampleyear), distance = "jaccard")  ## need to change variable name at somepoint to avoid confusion...

anova_terms <- anova(rda_index, by = "terms", permutations = 999)
anova_terms 

summary(rda_index)

# Assuming capscale_model is already defined
eigenvalues <- rda_index$CA$eig
total_inertia <- sum(eigenvalues)

# Calculate proportion explained for CAP1 and CAP2
proportion_variance <- eigenvalues / total_inertia * 100

# First use the regular plot option to get a feel for the data
plot(rda_index, scaling = 2, choices = c(1,2))
#plot(rda_index, scaling = 2, choices = c(1,3))


# Extract site and species scores
site_scores <- as.data.frame(vegan::scores(rda_index, display = "sites", choices = 1:3))
#site_scores <- as.data.frame(rda_index$CCA$u[, 1:3])
site_scores$Sample <- rownames(site_scores)  # Add sample names

species_scores <- as.data.frame(vegan::scores(rda_index, display = "species", choices = 1:3))
#species_scores <- as.data.frame(rda_index$CCA$v[, 1:3])
species_scores$Species <- rownames(species_scores)  # Add species names

# Extract biplot arrows (for explanatory variables)
biplot_arrows <- as.data.frame(vegan::scores(rda_index, display = "bp", choices = 1:3))
#biplot_arrows <- as.data.frame(rda_index$CCA$biplot[, 1:3])
biplot_arrows$variable <- rownames(biplot_arrows)

biplot_arrows <- biplot_arrows %>%
  mutate(variable = ifelse(variable == "fish_meta$eq_latitude", "latitude", variable),
         #variable = ifelse(variable == "meta_for_nmds$depth", "depth", variable),
         variable = ifelse(variable == "as.factor(fish_meta$sampleyear)2013", "year_2013", variable),
         variable = ifelse(variable == "as.factor(fish_meta$sampleyear)2017", "year_2017", variable),
         variable = ifelse(variable == "as.factor(fish_meta$sampleyear)2019", "year_2019", variable)) #,
         #variable = ifelse(variable == "meta_for_nmds$temp_C", "temperature", variable),
         #variable = ifelse(variable == "meta_for_nmds$salinity", "salinity", variable))

## Plot CAP1 vs CAP2 - color sites by latitude
plot_lat <- ggplot() +
  # Plot site scores
  geom_point(data = site_scores, aes(x = CAP1, y = CAP2, color = fish_meta$eq_latitude), size = 3) +
  #geom_text_repel(data = site_scores, aes(x = CAP1, y = CAP2, label = Sample), size = 4) +
  scale_color_viridis(option = "cividis",  direction = -1) +
  
  # Plot species scores
  geom_point(data = species_scores, aes(x = CAP1, y = CAP2), color = "red", size = 2) +
  geom_text_repel(data = species_scores, aes(x = CAP1, y = CAP2, label = Species), color = "red", size = 2) +
  
  # Plot arrows
  geom_segment(data = biplot_arrows, aes(x = 0, y = 0, xend = CAP1, yend = CAP2), 
               arrow = arrow(length = unit(0.3, "cm")), color = "black") + # Plot biplot arrows
  geom_text(data = biplot_arrows, aes(x = CAP1 * 1.1, y = CAP2 * 1.1, label = variable), color = "black") + # Label arrows
  
  # Add labels and theme
  labs(title = "CAP Plot", 
       x = paste("CAP1 (", round(proportion_variance[1], 2), "% variance explained)", sep = ""), 
       y = paste("CAP2 (", round(proportion_variance[2], 2), "% variance explained)", sep = ""),
       color = "latitude") +
  theme_minimal() #+
  #theme(legend.position = "none")


## Plot CAP1 vs CAP2 - color sites by year
plot_year <- ggplot() +
  # Plot site scores
  geom_point(data = site_scores, aes(x = CAP1, y = CAP2, color = as.factor(fish_meta$sampleyear)), size = 3) +
  #geom_text_repel(data = site_scores, aes(x = CAP1, y = CAP2, label = Sample), size = 4) +
  
  # Plot species scores
  geom_point(data = species_scores, aes(x = CAP1, y = CAP2), color = "black", size = 2) +
  geom_text_repel(data = species_scores, aes(x = CAP1, y = CAP2, label = Species), color = "black", size = 2) +
  
  # Plot arrows
  geom_segment(data = biplot_arrows, aes(x = 0, y = 0, xend = CAP1, yend = CAP2), 
               arrow = arrow(length = unit(0.3, "cm")), color = "black") + # Plot biplot arrows
  geom_text(data = biplot_arrows, aes(x = CAP1 * 1.1, y = CAP2 * 1.1, label = variable), color = "black") + # Label arrows
  
  # Add labels and theme
  labs(title = "CAP Plot", 
       x = paste("CAP1 (", round(proportion_variance[1], 2), "% variance explained)", sep = ""), 
       y = paste("CAP2 (", round(proportion_variance[2], 2), "% variance explained)", sep = ""),  
       color = "year") +
  theme_minimal() #+
  #theme(legend.position = "none")

plot_lat
plot_year
```


hierarchial cluster analysis 
```{r}
#let's just go with one year of data for now...
sub_2013 <- fish_wide_filtered %>%
  filter(sampleyear == 2013)

meta_2013 <- sub_2013[,c(1:7)]
fish_2013 <- sub_2013[,-c(1:7)]

dist_2013 <- vegdist(fish_2013, method = "jaccard")

# Cluster Dendrogram
hc <- hclust(dist_2013, method = "ward.D2")
plot(hc, labels = meta_2013$stationid)

#groups <- cutree(hc, k = 2)   ## just starting with k=3
#adonis2(dist_2013 ~ groups)

# Number of clusters to evaluate
k.max <- 10
wss <- numeric(k.max)

# Loop through k values to calculate WSS
for (k in 1:k.max) {
  # Cut the dendrogram into k groups
  cluster_assignment <- cutree(hc, k)
  
  # Calculate total within-cluster sum of squares
  wss[k] <- sum(sapply(unique(cluster_assignment), function(cluster) {
    sum((dist(fish_2013[cluster_assignment == cluster, ])^2))  # Sum of squares for each cluster
  }))
}

# Plot the elbow curve
plot(1:k.max, wss, type = "b", pch = 19, frame = FALSE, 
     xlab = "Number of Clusters (k)", 
     ylab = "Total Within-Cluster Sum of Squares (WSS)",
     main = "Elbow Method for Optimal k")

groups <- cutree(hc, k = 4)   ## just starting with k=3
adonis2(dist_2013 ~ groups)

plot(hc, labels = meta_2013$stationid)
rect.hclust(hc, k = 4, border = "red")  # You can specify color here


#library(ggdendro)
#ggdendrogram(hc)

library(dendextend)
# Create a dendrogram object
dend <- as.dendrogram(hc)
# Create a color palette (e.g., blue to red gradient based on latitude)
latitudes <- meta_2013$eq_latitude
color_palette <- colorRampPalette(c("blue", "red"))(length(latitudes))
# Map latitudes to colors
latitude_colors <- color_palette[rank(latitudes)]  # Rank the latitudes and assign colors
# Assign these colors to the labels (tips) of the dendrogram
labels_colors(dend) <- latitude_colors
# Plot the dendrogram with colored tips based on latitude
plot(dend, main = "Dendrogram with Tips Colored by Latitude")
rect.hclust(hc, k = 4, border = "red")  # You can specify color here

```

okay, not all groups are predicted by latitude 

run similarity percentages (SIMPER) analysis using vegan
```{r}
simper_out <- simper(fish_2013, group = groups)
simper_out

temp <- summary(simper_out)

# take output out into a single dataframe 
# Initialize an empty data frame
combined_df <- data.frame()
# Loop through the list to combine data frames
for (group_name in names(simper_out)) {
  # Extract the current data frame
  temp_df <- simper_out[[group_name]]
  
  # Add a new column with the group name
  temp_df$Group <- group_name
  
  # Combine with the main data frame
  combined_df <- bind_rows(combined_df, temp_df)
}

# what species have p < 0.05
simper_sign <- combined_df %>%
  filter(average > 0) %>%
  filter(p <= 0.05)
```

arctic cod and herring are driving station clusters 

now i want to plot those cluster onto a map to visualize any location patterns 
```{r}
meta_2013$group <- groups
```

load libraries
```{r}
library(rnaturalearth)
library(rnaturalearthdata)
world <- ne_countries(scale = "medium", returnclass = "sf")
```

```{r}
min_lat <- min(meta_2013$eq_latitude, na.rm = T)
max_lat <- max(meta_2013$eq_latitude, na.rm = T)

min_lon <- min(meta_2013$eq_longitude, na.rm = T)
max_lon <- max(meta_2013$eq_longitude, na.rm = T)

meta_2013$group <- as.factor(meta_2013$group)
  
ggplot(data = world) +
  geom_sf() +
  geom_point(data = meta_2013, aes(x=eq_longitude, y=eq_latitude, color = group)) +
  #scale_color_gradient(low = "blue", high = "red", na.value = "black") +  # Color gradient for counts
  #facet_grid(collection_year~CommonName) + 
  coord_sf(xlim = c(min_lon-2, max_lon+2), ylim = c(min_lat-1, max_lat+1), expand = FALSE) +
  theme_bw() +xlab("Longitude") +ylab("Latitude")+
  labs(color = "count") +
  theme(axis.text.x = element_text(angle = 90))
```


indicator species
```{r}
library(labdsv)

# Perform indicator species analysis
indicator_year <- indval(fish_binary, fish_meta$sampleyear, perm = 999)
summary(indicator_year)
```

```{r}
# this only works for categorical variables, so let me make some dummy groups for this 
binned_lat <- cut(fish_meta$eq_latitude, breaks = quantile(fish_meta$eq_latitude, probs = seq(0,1,0.25)), include.lowest = TRUE)

#meta_for_nmds <- meta_for_nmds %>%
#  mutate(latitude_group = ifelse(latitude <= 65, "low", NA),
#         latitude_group = ifelse(latitude > 65 & latitude < 71, "middle", latitude_group),
#         latitude_group = ifelse(latitude >= 71, "high", latitude_group))

#indicator_latitude<- indval(index_df_wide[,-1], meta_for_nmds$latitude_group, perm = 999)
indicator_latitude<- indval(fish_binary, binned_lat, perm = 999)
indicator_latitude
summary(indicator_latitude)
```

taxa richness (fish only) by year and latitude
```{r}
taxon_table_with_year_and_lat <- cbind(fish_meta$sampleyear, fish_meta$eq_latitude, fish_binary)
colnames(taxon_table_with_year_and_lat)[1] <- "year"  # Rename the year column
colnames(taxon_table_with_year_and_lat)[2] <- "latitude"
#colnames(taxon_table_with_year_and_lat)[3] <- "depth"
#colnames(taxon_table_with_year_and_lat)[4] <- "temp"

# Exclude the "year" column and calculate species richness using rowSums
taxon_table_with_year_and_lat <- taxon_table_with_year_and_lat %>%
  mutate(richness = rowSums(select(., -c(year,latitude)) > 0))

ggplot(taxon_table_with_year_and_lat, aes(x = latitude, y = richness, color = as.factor(year))) +
  geom_point() +   # Scatter plot points
  geom_smooth(method = "loess", se = TRUE) +  # Smoothed line (loess curve)
  labs(x = "Latitude", y = "Taxa Richness", title = "Species Richness by Latitude", color = "latitude") +
  theme_minimal()                 # Apply a clean minimal theme

```

much higher spp richness ~70N in 2017 than in other years 
- need to check how taxa were id'd - ie make sure level of classifications is consistent over the years

check out what's going on with arctic cod and pollock by latitude 
disclaimer:  i don't believe this version of the data has been corrected for the mis-id'd cod

```{r}
fish_wide_all_stations <- fish_wide_all_stations %>%
  mutate(lat_bin = ifelse(eq_latitude <= 64, "<64", NA),
         lat_bin = ifelse(eq_latitude > 64 & eq_latitude <= 66, "64-66", lat_bin),
         lat_bin = ifelse(eq_latitude > 66 & eq_latitude <= 68, "66-68", lat_bin),
         lat_bin = ifelse(eq_latitude > 68 & eq_latitude <= 70, "68-70", lat_bin),
         lat_bin = ifelse(eq_latitude > 70 & eq_latitude <= 72, "70-72", lat_bin),
         lat_bin = ifelse(eq_latitude > 72, ">72", lat_bin))

haul_summary <- fish_wide_all_stations %>%
  group_by(sampleyear, lat_bin) %>%
  summarise(unique_hauls = n(),
            AC = sum(`Boreogadus saida` > 0),
            WP = sum(`Gadus chalcogrammus` > 0),
            prop_AC = AC/unique_hauls,
            prop_WP = WP/unique_hauls)

my_order <- c("<64", "64-66", "66-68", "68-70", "70-72", ">72")

haul_summary$lat_bin <- factor(haul_summary$lat_bin, levels = my_order)

df_prop_w_pos <- haul_summary %>%
  select(sampleyear, lat_bin, prop_AC, prop_WP) %>% #  , prop_PC, prop_SC, prop_PolC) %>%
  pivot_longer(cols = c(3:4), values_to = "prop", names_to = "taxon") %>%
  mutate(CommonName = ifelse(taxon == "prop_AC", "Arctic cod", NA),
         CommonName = ifelse(taxon == "prop_WP", "Walleye pollock", CommonName)) #,
         #CommonName = ifelse(taxon == "prop_PC", "Pacific cod", CommonName),
         #CommonName = ifelse(taxon == "prop_SC", "Saffron cod", CommonName),
         #CommonName = ifelse(taxon == "prop_PolC", "Polar cod", CommonName))

df_prop_w_pos %>%
  ggplot(aes(x=CommonName, y=lat_bin, fill = prop)) +
  geom_tile() + 
  scale_fill_gradient(low = "lightblue", high = "darkblue") +  # Adjust color scale as needed
  #geom_text(data = samp_size, aes(x=lat_bin, y = total_samples + 2, label = total_samples)) +
  facet_grid(~sampleyear) + 
  labs(title = "Heatmap of Proportion of Hauls with Detection",
       x = "Species",
       y = "Latitude",
       fill = "Proportion of Hauls with Detection") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability  

```

okay, general trends are there but the data is very messy. 
