---
title: "testing code..."
output: html_document
date: "2024-10-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

try out accumulation curve to evaluate sequencing depth 
```{r}
library(iNEXT)
library(ggplot2)

#iNext crashes easily because of too many samples.  will need to break this up by project and year and subsets of samples...
DBO_2021 <- taxon_table_1 %>%
filter(collection_year == 2021)

# need N lists of species read counts 

taxon_table_2 <- DBO_2021 %>%
   pivot_wider(names_from = taxon, values_from = tot_reads)

# now convert the taxon table to lists of species abundances (one list per pcr rep)

# make a variable for the first and last ASV column in the table
tax_first <- which(colnames(taxon_table_2) == "Acantholumpenus mackayi")  
tax_last <- ncol(taxon_table_2)

# okay, need to go back to long format 
taxon_long <- taxon_table_2 %>%
  pivot_longer(cols = tax_first:tax_last, names_to = "taxon", values_to = "reads") %>%    
  filter(reads > 0) %>% # filter out entries with no reads
  arrange(extraction_ID, desc(reads))  # sort reads in descending order and group by sample

#st <- taxon_long %>%
#  filter(location1 %in% c("5"))

# Nest the data to create a list of species reads for each sample
read_lists <- taxon_long %>%
  #group_by(sample_ID_date) %>%
  group_by(extraction_ID) %>%
  nest() %>%
  mutate(
    species_read_list = map(data, ~ .x %>% select(taxon, reads))
  ) %>%
  #select(sample_ID_date, species_read_list)
  select(extraction_ID, species_read_list)

#Convert to a list of lists
read_list_final <- read_lists %>%
  pull(species_read_list) %>%
  #setNames(read_lists$sample_ID_date)
  setNames(read_lists$extraction_ID)

# Convert each table to a numeric vector of abundances
read_lists_vector <- map(read_list_final, ~ {
  .x %>%
    filter(reads > 0) %>%
    arrange(desc(reads)) %>%
    pull(reads)
})

# knots is the number of intervals
# endpoint is the max read count 
# m can designate sample size 

# try to run iNEXT - the function doesn't like non-integers so lots of warning will pop up.
#out_nbs2023_st38 <- iNEXT(read_lists_vector, q = 0, datatype = "abundance")
#out_nbs2023_st2_80knots <- iNEXT(read_lists_vector, q = 0, datatype = "abundance", knots = 80, endpoint = 100000) 
#out_nbs2021_st46 <- iNEXT(read_lists_vector, q = 0, datatype = "abundance", knots = 80)

#set a series of sample sizes for (m) 
m <- c(1,50,100,250,500,1000,2000,3000,4000,5000,10000,20000,30000,40000)

out_dbo21_m <- iNEXT(read_lists_vector, q = 0, datatype = "abundance", size = m)
#out_nbs2021_st5_m <- iNEXT(read_lists_vector, q = 0, datatype = "abundance", size = m)

```

check out the output 
```{r}
out_nbs2021_st20_m$DataInfo
out_nbs2021_st20_m$iNextEst
out_nbs2021_st20_m$AsyEst
```
 
custom plot since ggiNEXT is giving me color problems 
```{r}
df <- fortify(out_nbs2021_st5_m, type=1)
df.point <- df[which(df$Method=="Observed"),]
df.line <- df[which(df$Method!="Observed"),]
df.line$Method <- factor(df.line$Method,  c("Rarefaction", "Extrapolation"), c("Rarefaction", "Extrapolation"))

my_plot <- ggplot(df, aes(x=x, y=y, colour=Assemblage)) +  
  geom_point(size=5, data=df.point) + 
  geom_line(aes(linetype=Method), lwd=1.5, data=df.line) + 
  geom_ribbon(aes(ymin=y.lwr, ymax=y.upr, fill=Assemblage, colour=NULL), alpha=0.2) + 
  labs(x="number of sequencing reads per 1-L bottle", y="Number of detected taxa", 
       title = "NBS 2021 Station 5") + 
  theme(legend.position = "bottom",  legend.title=element_blank(), text=element_text(size=18), legend.box = "vertical") +
  xlim(0, 10000) + 
  theme_minimal()
my_plot
```
 
```{r}
#ggsave("/home/kimberly.ledger/BeringSea_Arctic_eDNA/outputs/accumulation_curves/NBS_2023_st12_23_38_43_zoom.png", plot = my_plot, dpi = 150, width = 8, height = 6)
```

```{r}
#extract iNEXT result 
accumulation_data <- out_nbs2021_st5_m$iNextEst[[1]] 
```

maybe try to come back see if i can quantitatively determine the number of reads to be within x % of the max/observed species richness???? 

this finds the number of species for a certain number of sampled sites - would likely need to break down sites by year and location to make this information useful for my project
```{r}
accum_curve <- specaccum(taxon_table_wide, "random")
summary(accum_curve)
plot(accum_curve)
```


make some plots for DBO 2021

spp in DBO 2021
```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2021") %>%
  filter(tot_reads > 0) %>%
  group_by(taxon) %>%
  summarize(n_reads = sum(tot_reads)) %>%
  arrange(desc(n_reads))
```
```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2021") %>%
  filter(location2 == "DBO_1") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 1 - 2021") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2021") %>%
  filter(location2 == "DBO_2") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 2 - 2021") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2021") %>%
  filter(location2 == "DBO_3") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 3 - 2021") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2021") %>%
  filter(location2 == "DBO_4") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 4 - 2021") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2021") %>%
  filter(location2 %in% c("CEO", "C2", "M8")) %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Moorings - 2021") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

make some plots for DBO 2023

spp in DBO 2023 
```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(tot_reads > 0) %>%
  group_by(taxon) %>%
  summarize(n_reads = sum(tot_reads)) %>%
  arrange(desc(n_reads))
```


```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_1") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 1 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r, fig.width=7.5, fig.height=5}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  #filter(location2 == "DBO_1") %>%
  filter(location1 == "DBO_1.4") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "relative read abundance (%)",
    x = "depth",
    title = "DBO Station 1 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_1") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  group_by(taxon) %>%
  summarize(mean_prop = mean(prop)) %>%
  arrange(desc(mean_prop))
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_2") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 2 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r, fig.width=8.5, fig.height=5}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_2") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "relative read abundance (%)",
    x = "depth",
    title = "DBO Station 2 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_2") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  group_by(taxon) %>%
  summarize(mean_prop = mean(prop)) %>%
  arrange(desc(mean_prop))
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_3") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 3 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r, fig.width=8.5, fig.height=5}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_3") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "relative read abundance (%)",
    x = "depth",
    title = "DBO Station 3 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```


```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_3") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  group_by(taxon) %>%
  summarize(mean_prop = mean(prop)) %>%
  arrange(desc(mean_prop))
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_4") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 4 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```


```{r, fig.width=8.5, fig.height=5}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_4") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "relative read abundance (%)",
    x = "depth",
    title = "DBO Station 4 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_4") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  group_by(taxon) %>%
  summarize(mean_prop = mean(prop)) %>%
  arrange(desc(mean_prop))
```


```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_5") %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Station 5 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r, fig.width=8.5, fig.height=5}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_5") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "relative read abundance (%)",
    x = "depth",
    title = "DBO Station 5 - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 == "DBO_5") %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  group_by(taxon) %>%
  summarize(mean_prop = mean(prop)) %>%
  arrange(desc(mean_prop))
```

```{r}
taxon_table_meta %>%
  filter(project_year == "DBO_2023") %>%
  filter(location2 %in% c("BF2", "CK9", "M8")) %>%
  filter(tot_reads > 0) %>%
  #ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~location1, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO Moorings - 2023") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```



now let's plot relative read abundance 
```{r}
taxon_table_meta %>%
  unite(col = "station_depth", location1, depth, sep = "_", remove = F) %>%
  filter(project_year == "NBS_2021") %>%
  filter(location1 %in% c("2", "8", "11", "17")) %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  ggplot(aes(x=extraction_ID, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(~station_depth, scales = "free_x") + 
  labs(
    y = "relative read abundance (%)",
    x = "sample ID",
    title = "NBS 2021") + 
  theme(
    axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```



is there a clear latitude cutoff/change point in species comps? 

```{r}
meta %>%
  filter(latitude > 70) %>%
  select(location2) %>%
  unique()

meta %>%
  filter(latitude < 70) %>%
  select(location2) %>%
  unique()
```

```{r}
taxon_table_meta %>%
  filter(project == "DBO") %>%
  filter(latitude > 70) %>%
  filter(tot_reads > 0) %>%
  ggplot(aes(x=extraction_ID, y=tot_reads, fill=taxon)) +
  #ggplot(aes(x=depth, y=tot_reads, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(collection_year~location2, scales = "free_x") + 
  labs(
    y = "number of sequencing reads",
    x = "depth",
    title = "DBO above 70N") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r, fig.width=8.5, fig.height=5}
taxon_table_meta %>%
  filter(project == "DBO") %>%
  filter(latitude > 70) %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  ggplot(aes(x=extraction_ID, y=prop, fill=taxon)) +
  #ggplot(aes(x=depth, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(collection_year~location2, scales = "free_x") + 
  labs(
    y = "relative read abundance (%)",
    x = "depth",
    title = "DBO above 70N") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project == "DBO") %>%
  filter(latitude > 70) %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  group_by(taxon) %>%
  summarize(mean_prop = mean(prop)) %>%
  arrange(desc(mean_prop))

taxon_table_meta %>%
  filter(project == "DBO") %>%
  filter(latitude > 70) %>%
  filter(tot_reads > 0) %>%
  group_by(taxon) %>%
  summarize(total_reads = mean(tot_reads)) %>%
  arrange(desc(total_reads))
```




```{r, fig.width=8.5, fig.height=5}
taxon_table_meta %>%
  filter(project == "DBO") %>%
  filter(latitude < 70) %>%
  filter(tot_reads > 0) %>%
  group_by(extraction_ID) %>%
  mutate(sum=sum(tot_reads)) %>%
  mutate(prop = tot_reads/sum) %>%
  ggplot(aes(x=extraction_ID, y=prop, fill=taxon)) +
  #ggplot(aes(x=depth, y=prop, fill=taxon)) +
  geom_bar(stat = "identity") + 
  theme_bw() +
  facet_grid(collection_year~location2, scales = "free_x") + 
  labs(
    y = "relative read abundance (%)",
    x = "depth",
    title = "DBO belove 70N") + 
  theme(
    #axis.text.x = element_text(angle = 90, hjust = 0.95),
    legend.text = element_text(size = 8),
    legend.key.size = unit(0.3, "cm"),
    legend.position = "bottom",
    legend.title = element_blank()
  )
```

```{r}
taxon_table_meta %>%
  filter(project == "DBO") %>%
  filter(latitude < 70) %>%
  filter(tot_reads > 0) %>%
  group_by(taxon) %>%
  summarize(total_reads = mean(tot_reads)) %>%
  arrange(desc(total_reads))
```






## some extra code... 


```{r}
# dbo_pres_21 <- dbo_pres %>%
#   filter(collection_year == 2021)
# 
# matrix_21 <- dbo_pres_21[,-c(1:4)]
# meta_21 <- dbo_pres_21[,c(1:4)]
# 
# dbo_pres_23 <- dbo_pres %>%
#   filter(collection_year == 2023)
# 
# matrix_23 <- dbo_pres_23[,-c(1:4)]
# meta_23 <- dbo_pres_23[,c(1:4)]

matrix <- dbo_pres[,-c(1:6)]
meta <- dbo_pres[,c(1:6)]
```



```{r}
#NMS_data<-metaMDS(matrix_21, distance = "jaccard", k=2, try=100, autotransform = TRUE, maxit=1000)
#NMS_data<-metaMDS(matrix_21, distance = "bray", k=2, try=100, autotransform = TRUE, maxit=1000)

#NMS_data<-metaMDS(matrix_23, distance = "bray", k=2, try=100, autotransform = TRUE, maxit=1000)

#NMS_data<-metaMDS(matrix, distance = "jaccard", k=2, try=100, autotransform = TRUE, maxit=1000)
NMS_data<-metaMDS(matrix, distance = "bray", k=2, try=100, autotransform = TRUE, maxit=1000)
```

